A Spiking Independent Accumulator Model for Winner-Take-All Computation

### 2017-05-04 doc/paper/paper.tex
Minor updates to paper.

Note that the layer 2 decoded output of a single choice shows more variance in the IA network (Fig.~\ref{fig:indacc}), but the separation of the output for different choices is more relevant in interpreting the output.

@@

\section{Results}
We find that the ability to form a clear decision of the LCA network decreases with more noise and less target separation  (see Fig.~\ref{fig:decisions}).
Also, the magnitude of the inputs has an important influence.
For a small inputs with $u=0.2$ the winner will mostly not exceed the 0.15 threshold with noise present.
The best performance is achieved with medium inputs with $u=0.6$.
Higher inputs make it more likely that runner-ups will exceed the 0.15 threshold, especially with a small target separation.
In contrast to the LCA network, the IA network manages to form a clear decision in every trial (not explicitly shown in Figure~\ref{fig:decisions}, but all data points fall on the grey line).
\begin{figure*}[t]
    \centering
    \includegraphics[scale=0.9]{figures/decisions}
    \caption{
        Fraction of trials with a clear decision for the LCA network (which also exactly matches the fraction of correct trials).
        Each plot shows data for a different input magnitude $u$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum, which coincides with the performance of the IA network.
    }\label{fig:decisions}
\end{figure*}

Interestingly, for all clear decisions the correct winner was determined by the LCA network.
Thus, a plot of correct trials looks identical to Figure~\ref{fig:decisions}, with slightly different error bars.
While always reaching a clear decision, the decisions of the IA network are not always correct.
Overall, the IA performance tends to be worse than the LCA performance for a high input magnitude, but better for smaller inputs (Fig.~\ref{fig:correct}a,\,b).
We can greatly improve the IA performance by increasing $\tau_1$ to \SI{0.5}{\second} which slows down the integration of evidence (Fig.~\ref{fig:correct}c).
This improves the IA performance to be above the LCA performance for high baselines, but it will also increase the decisions times of the IA network.
For a low baseline with $u=0.2$, it makes the IA network unable to decide within the allocated time frame, but given more time it would still reach a decision.
%A further improvement to the IA network is possible by eliminating a systematic bias introduced in the second layer of the IA network.
%In the NEF the neuron gain and bias parameters are usually picked from random distributions.
%As such the layer 2 populations will become active at slightly different thresholds.
%By using identical layer 2 populations, i.e., identical gain and bias parameters for each population, this bias can be eliminated and the IA network performs even better (Fig.~\ref{fig:correct}d).
\begin{figure*}[t]
    \centering
    \includegraphics[scale=0.9]{figures/correct}
    \caption{
        Fraction of correct trials for the IA network.
        Each plot shows data for a different combination of input magnitude $u$ and integration time-constant $\tau_1$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum.
    }\label{fig:correct}
\end{figure*}

@@

Finally, looking at the transient responses indicates that both models might produce outputs of losing choices.
For the LCA network the magnitude of the transient response mainly increases with the amount of noise (Fig.~\ref{fig:transient}a).
For the IA network, transient outputs are smaller in noisy conditions, but can be higher than for the LCA network in less noisy conditions with small target separations.
The magnitude of such transient responses is reduced by adjusting $\tau_1$ to 0.5, at the cost of a slower decision.
\begin{figure*}
    \centering
    \includegraphics[scale=0.9]{figures/transient}
    \caption{
        Transient response (highest output of losing choices) for the LCA and IA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:transient}
\end{figure*}

@@

In contrast, the IA network is better suited for situations where a discrete sequence of decisions is required.
After selecting a winner, the model's decision will persist due to self-excitation, even in the absence of input.
Thus, after making a decision, it is necessary to reset the model by inhibiting the winning accumulator.
This limits how quickly successive decisions can be made and reduces the ability to react to changing inputs.
However, once a decision is made, the network provides a stable output.
As an example, we intend to use the IA network in a free recall memory model where a clear selection of successive responses needs to be made, but correctly identifying the single item remembered most strongly is not required.

Both models might produce a transient response that may be interpreted as a decision, which can make the detection of decisions problematic.
This is of special relevance when incorporating the networks into larger cognitive models.
For the LCA network, the transient behaviour is inherent to its design; there will always be an initial rise of all state variables (that receive input) before the mutual inhibition grows strong enough to push them back to zero.
In the IA network, however, such transient responses may be avoided by choosing appropriate $\tau_1$ and $\vartheta$.
It should be noted that other recurrently connected readout mechanisms have been proposed for the two choice case that produce bursting outputs somewhat comparable to the IA network~\cite{lo2006}.
However, the evidence integration uses competition by means of pooled inhibition makes it more similar to the LCA than the IA network and the behaviour under highly noisy conditions is unknown.

@@

We have not yet investigated the agreement of these mechanisms with neurobiological and behavioural data, although this has been done before for other WTA networks~\cite{gold2007,smith2004}.
The implementation in spiking neurons, however, provides some basic biological plausibility and more readily permits comparisons with neural data.
In particular, the IA network predicts that the firing rates for neurons in the first layer will rise up to a threshold, and that neurons in the second layer will not become active or inhibit the first layer until reaching this threshold.
Neurons in the macaque lateral intraparietal area give a similar step response~\cite{latimer2015}.
In contrast, the LCA network predicts that the firing of any neurons will proportionately inhibit all other neurons that do not represent the same state.
With regard to behavioural data, different effects for decision times are predicted as the number of choices increases:
for the LCA network decisions become slower due to the mutual interaction, but for the IA network decisions will take less time because only a single accumulator needs to exceed the threshold.
Nevertheless, evidence for one network does not exclude the possibility that the other network is employed for different tasks or by different brain areas.
Relatedly, it might be more plausible to distribute the representation of all state variables over a single population of neurons.
This is directly supported by the NEF, but we chose to leave this to future work to keep the current analysis free from potential interactions of the state variables introduced by such a distributed representation.

>>>>

Note that the decoded output from layer 2 of the IA network has higher variance than the LCA network (Fig.~\ref{fig:indacc}), but the separation of the output for different choices is more relevant than the variance in interpreting the output.

@@

\section{Results}
We find that the ability to form a clear decision of the LCA network decreases with more noise and less target separation  (see Fig.~\ref{fig:decisions}).
Also, the magnitude of the inputs has an important influence.
For a small inputs with $u=0.2$ the winner will mostly not exceed the 0.15 threshold with noise present.
The best performance is achieved with medium inputs with $u=0.6$.
Higher inputs make it more likely that runner-ups will exceed the 0.15 threshold, especially with a small target separation.
In contrast to the LCA network, the IA network manages to form a clear decision in every trial (not explicitly shown in Figure~\ref{fig:decisions}, but all data points fall on the grey line).
\begin{figure*}[t]
    \centering
    \includegraphics[scale=0.95]{figures/decisions}
    \caption{
        Fraction of trials with a clear decision for the LCA network (which also exactly matches the fraction of correct trials).
        Each plot shows data for a different input magnitude $u$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum, which coincides with the performance of the IA network.
    }\label{fig:decisions}
\end{figure*}

Interestingly, for all clear decisions the correct winner was determined by the LCA network.
Thus, a plot of correct trials looks identical to Figure~\ref{fig:decisions}, with slightly different error bars.
While always reaching a clear decision, the decisions of the IA network are not always correct.
Overall, the IA performance tends to be worse than the LCA performance for a high input magnitude, but better for smaller inputs (Fig.~\ref{fig:correct}a,\,b).
We can greatly improve the IA performance by increasing $\tau_1$ to \SI{0.5}{\second} which slows down the integration of evidence (Fig.~\ref{fig:correct}c).
This improves the IA performance to be above the LCA performance for high baselines, but it will also increase the decisions times of the IA network.
For a low baseline with $u=0.2$, it makes the IA network unable to decide within the allocated time frame, but given more time it would still reach a decision.
%A further improvement to the IA network is possible by eliminating a systematic bias introduced in the second layer of the IA network.
%In the NEF the neuron gain and bias parameters are usually picked from random distributions.
%As such the layer 2 populations will become active at slightly different thresholds.
%By using identical layer 2 populations, i.e., identical gain and bias parameters for each population, this bias can be eliminated and the IA network performs even better (Fig.~\ref{fig:correct}d).
\begin{figure*}[t]
    \centering
    \includegraphics[scale=0.95]{figures/correct}
    \caption{
        Fraction of correct trials for the IA network.
        Each plot shows data for a different combination of input magnitude $u$ and integration time-constant $\tau_1$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum.
    }\label{fig:correct}
\end{figure*}

@@

Finally, looking at the transient responses indicates that both models might produce outputs of losing choices.
For the LCA network the magnitude of the transient response mainly increases with the amount of noise (Fig.~\ref{fig:transient}a).
For the IA network, transient outputs are smaller in noisy conditions, but can be higher than for the LCA network in less noisy conditions with small target separations.
The magnitude of such transient responses is reduced by adjusting $\tau_1$ to 0.5, at the cost of a slower decision.
\begin{figure*}
    \centering
    \includegraphics[scale=0.92]{figures/transient}
    \caption{
        Transient response (highest output of losing choices) for the LCA and IA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:transient}
\end{figure*}

@@

In contrast, the IA network is better suited for situations where a discrete sequence of decisions is required.
After selecting a winner, the model's decision will persist due to self-excitation, even in the absence of input.
Thus, after making a decision, it is necessary to reset the model by inhibiting the winning accumulator.
This limits how quickly successive decisions can be made and reduces the ability to react to changing inputs.
However, once a decision is made, the network provides a stable output.
For example, we intend to use the IA network in a model of free recall to output a sequence of WTA responses.
In this case, the model requires stable recall in the presence of noise, even if each response may not be the ``true'' winner.

Both models might produce a transient response that may be interpreted as a decision, which can make the detection of decisions problematic.
This is of special relevance when incorporating the networks into larger cognitive models.
For the LCA network, the transient behaviour is inherent to its design; there will always be an initial rise of all state variables (that receive input) before the mutual inhibition grows strong enough to push them back to zero.
In the IA network, however, such transient responses may be avoided by choosing appropriate $\tau_1$ and $\vartheta$.
It should be noted that other recurrently connected readout mechanisms have been proposed for the two choice case that produce bursting outputs somewhat comparable to the IA network~\cite{lo2006}.
However, the evidence integration in the \citeA{lo2006} model uses competition via pooled inhibition, making it more similar to the LCA then the IA network.

@@

We have not yet investigated the agreement of these mechanisms with neurobiological and behavioural data, although this has been done before for other WTA networks~\cite{gold2007,smith2004}.
The implementation in spiking neurons, however, provides some basic biological plausibility and more readily permits comparisons with neural data.
In particular, the IA network predicts that the firing rates for neurons in the first layer will rise up to a threshold, and that neurons in the second layer will not become active or inhibit the first layer until reaching this threshold.
Neurons in the macaque lateral intraparietal area exhibit a similar step response~\cite{latimer2015}.
In contrast, the LCA network predicts that the firing of any neurons will proportionately inhibit all other neurons that do not represent the same state.
With regard to behavioural data, different effects for decision times are predicted as the number of choices increases:
for the LCA network decisions become slower due to the mutual interaction, but for the IA network decisions will take less time because only a single accumulator needs to exceed the threshold.
Nevertheless, evidence for one network does not exclude the possibility that the other network is employed for different tasks or by different brain areas.
Relatedly, it might be more plausible to distribute the representation of all state variables over a single population of neurons.
This is directly supported by the NEF, but we chose to leave this to future work to keep the current analysis free from potential interactions of the state variables introduced by such a distributed representation.
